Distributional Trees and Forests for Circular Data

### 2019-10-16 pkg/circtree/inst/wind_paper/draft/draft.Rnw
added reference on circular CRPS, information on subset and exponential weights in persistence model; added further fixmes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2795 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0a63e551fbc3b892f5ebd7e2cb4e732580afe6cf

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).}
%TODO: refer to data, Innsbruck airport
\label{fig:density_vM} 
\end{figure}

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
parameters of the von Mises distribution, however, in an iterative optimization procedure
requiring well fitting initial values for the parameters that are recommended to be calculated 
in advance in two separate models. Using the obtained initial value the model parameters are 
estimated and improved stepwise until convergence. While providing a well elaborated theory 
this methodology and the contained optimization problem have shown to be very complex and 
bear the risk of not converging if not applied thoughtfully \textit{FIXME: citation}. %\citealp{}.
Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

%% TODO: available approaches to wind direction forecasting

@@

Next to the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, nodes 7 and 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 12, 14 and 15). 
Node~11 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
increases clockwise from North ($0$ degree).

@@

\subsection{Data}
For both airports observations of various meteorological quantities measured
either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current wind direction, temperature, air pressure, humidity, ... at each 
of the weather stations also differences in these quantities between the considered 
stations are included. For example, differences in air pressure between Innsbruck and 
either a western or eastern location are expected to have a high influence on wind 
directions. Moreover, changes of these quantities over time are included as well as 
daytime. In total, 550 covariates are provided for Innsbruck and 522 covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) is calculated and summed up over each observation contained in
the learning data and the corresponding predicted distribution parameters of each of the models.

@@

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.
% TODO: references
To establish a distributional model of persistence for this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the exact same time points of the previous six days are considered. In order to account
for a stronger influence of observations closer to the prediction time point exponentially
increasing weights are included with the highest weight being assigned to the time point of the
previous day. For this selected set of observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.
% TODO: references
Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and the same day 
as well as the 30 surrounding days from all other years contained in the learning data are selected and
parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized Linear Model}
Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
on covariates by a linear combination of selected covariates connected to the parameter(s) via
a specified link function. The toolbox of generalized linear models ...
% TODO: references

>>>>

\begin{figure}[t]
\centering
\includegraphics[width=4cm]{density_vM.pdf}
\caption{Empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand).
\fixme{refer to data, Innsbruck airport}
}

\label{fig:density_vM} 
\end{figure}

@@

\section{Regression Trees and Forests for Circular Data}
\label{sec:tree_forest}
\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
For circular data the mixed model by \cite{Fisher:1993} follows the idea of modeling both
parameters of the von Mises distribution, however, in an iterative optimization procedure
requiring well fitting initial values for the parameters that are recommended to be calculated 
in advance in two separate models. Using the obtained initial value the model parameters are 
estimated and improved stepwise until convergence. While providing a well elaborated theory 
this methodology and the contained optimization problem have shown to be very complex and 
bear the risk of not converging if not applied thoughtfully. \fixme{citation} %\citealp{}.
Furthermore, these additive strategies work well for smooth effects but difficulties can arise 
in case of non-additive changes. Additionally, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\fixme{information on already available approaches to wind direction forecasting}

@@

Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the empirical (gray) 
and fitted von Mises (red) distribution of wind direction in each terminal node. Based on the 
fitted location parameters $\hat \mu$, the subgroups can be distinguished into the following 
wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, nodes 7 and 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 12, 14 and 15). 
Node~11 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.
Note that in the meteorological context wind direction is defined on the scale $[0,360]$ degree and 
increases clockwise from North ($0$ degree).

@@

\subsection{Evaluation Setting and Models Applied}
Along with distributional trees and forests for circular data three other models have been applied
in this study. In particular, they have been evaluated in a five-fold cross validation with each of 
the five years contained in the data set being used as testing data once after the models where 
learned on the remaining four years. 
In order to compare the performances of the employed models a circular version of the continuous 
ranked probability score (CRPS) as introduced by \cite{Grimit+Gneiting+Berrocal:2006} is calculated.
As the linear version of the CRPS it measures the difference between an observation
and the corresponding predicted distribution function in order to assess the probabilistic goodness 
of fit. In order to adjust this general methodology to circular data an additional auxiliary function
is employed which measures the angular distance between two circular variables. 
The resulting values are summed up over each observation contained in the learning data and the 
corresponding predicted distribution parameters of each of the models.

@@

\paragraph{Persistence}
Assuming that weather conditions will remain the same after a certain period of time is the
basic idea of persistence in meteorology. While this simple way of making predictions can
be very effective particularly under stable atmospheric circumstances it can fail remarkably
as soon as any changes take place, especially in case of rapid changes.
Therefore, it heavily depends on the specific features and the variability of the variable
of interest whether this approach can provide overall reliable predictions.

\fixme{references}

To establish a distributional model of persistence in this study a set of six observations
is considered to make a prediction for a selected day and time point. In particular,
the six previous time points are considered. In order to allow for a stronger influence of 
observations closer to the prediction time point exponential smoothing is employed, i.e.,
exponentially decaying weights are included with a smoothing factor of $0.5$ such that
in each prediction step the current prediction and the previous observation both have an 
equal influence rate of 50 percent. If a seventh or even earlier observations were considered 
as well their weights would be $<0.01$ and thus are not included anymore to simplify calculations.
For this selected set of weighted observations the distribution parameters of the von Mises 
distribution can be estimated as presented in Section~\ref{sec:prob_circ} such that a full
parametric model is specified allowing for probabilistic predictions and an evaluation of the CRPS.
As this approach only considers data from the previous six days a cross validation is not 
applicable, however, the same evaluation framework as for the other four models can be employed 
by obtaining predictions only for the testing data of the cross validation without using 
the corresponding learning data.

\paragraph{Climatology}
If there are no climatological changes within a certain number of years it can be expected that 
for a selected day of the year weather conditions will always be the same over all years 
considered.
Again, this idea provides a very simple and stable way of obtaining predictions for a considered 
meteorological variable of interest that can be very useful but its performance can also suffer 
dramatically from a high variance and abrupt changes, especially if they occur frequently during one day.

\fixme{add references}

Similar to the applied persistence model observations are selected to fit a distributional climatology
model. For a specified day and time point observations from the exact same time point and the same day 
as well as the 30 surrounding days from all other years contained in the learning data are selected and
parameters are estimated following again Section~\ref{sec:prob_circ}.

\paragraph{Generalized Linear Model}
Another commonly applied approach is to regress the parameter(s) of the von Mises distribution
on covariates by a linear combination of selected covariates connected to the parameter(s) via
a specified link function. The toolbox of generalized linear models ...

\fixme{add references}

\fixme{information on applied implementation}

\subsection{Data}
For both airports, Innsbruck and Vienna, observations of various meteorological quantities 
such as wind direction, temperature, air pressure and humidity
measured either directly at the airports or at surrounding weather stations are used as 
covariates in order to predict the wind direction at the corresponding airport 
in the following hour.
Next to the current measurements at each of the weather stations also differences in these 
quantities between the considered stations are included. For example, differences in air 
pressure between Innsbruck and either a western or eastern location are expected to have a 
high influence on wind directions. Moreover, changes of these quantities over time are 
included as well as daytime. In total, 550 covariates are provided for Innsbruck and 522 
covariates for Vienna. 
A list of all covariates together with their abbreviations as used in this paper
can be found in Table~\ref{tab:covariates}.
The data sets each provide observations from 5 years, in particular from 2014 to 2017, both included.
After eliminating observations with at least one missing value a total number of 26861 time points
for Innsbruck and 33853 for Vienna are available.
As winds with very low speed tend to vary their direction strongly they are not expected to be very 
useful/meaningful for predictions. Moreover, the performance and the stability of the employed reference 
models might suffer from the influence of these ``noise'' observations. Therefore, only observations
corresponding to a wind speed of at least $1$ m/s are considered.

\fixme{information on NAs}
